---
title: "Austin Nash's Notes for Spatial Capture Recapture"
format:
  html:
    embed-resources: true
theme: darkly
---

## Chapter 1
SCR is an extension to capture recapture methods, but also allows for the integration of a wide array of data types and ecological theory.
Common data streams used in SCR analyses include camera traps, DNA sampling, and acoustic recorders.
While typical data streams revolve around static detectors, search encounter methods such as detection dog surveys can be used in an SCR framework.
Encounter histories are commonly represented by a matrix with rows representing individuals and columns representing sampling events.
Authors present a motivating example of using traditional CR methods to estimate the density of bears. 
Going from N individuals to density requires strong assumptions, and density estimates vary widely based on the method selected. 
Traditional methods can't account for differences in exposure to detectors of individuals based on the location of their activity centers.
Traditional CR cannot accommodate trap level covariates. 
THe classical approach to estimating density involves adding a buffer to the MCP of the trap grid that is 0.5 times the size of the mean maximum distance moved. 
Another approach to correcting biases in traditional CR models is by using telemetry to estimate how often an animal temporarily emigrates from the trapping area. 
Another CR bias correction approach is using distance to edge of the trapping grid as an individual covariate on detectability, which is an intermediate step to full SCR.

Efford 2004 is the first formulation of a spatially explicit CR. 
This involved a point process model formulation and equated the observation process to a distance sampling problem. 
Observations of individuals represent a imperfectly observed, or thinned, point process. 
Another formalization is activity centers being parent nodes, with locations of individuals being child nodes, which is a poisson cluster process. 
Authors use the term home range and activity center interchangeably. 
SCR does not require the organism to be a central place forager, the activity center is essentially equivalent to a utilization distribution. 
The possible values of all activity center points are referred to as the state space. 
$$D = \frac {N}{A(S)} $$

## Chapter 2
Primarily review, covers the basics of statistical inference.
Review of probability notation, random variables, common distributions

Binomial Distribution - used in SCR as an encounter model for the number of times i individual is captured at trap j after k occasions. Probability of encounter is governed by the distance between the activity center and the detector. Also used as a prior on population size. 

Bernoulli Distribution - Binomial where n = 1, proper distribution to use when individuals have different detection probabilities (in the distance case have different detection kernels)

Multinomial distribution - commonly used in CR studies to model the encounter histories of individuals. 
Each category is each possible encounter history. 
The multinomial is used in scr to model when an individual can only be caught in a single trap in an ocassion (like for box traps, unlike camera traps), N = 1 and the cell probabillities are the probability of being captured in each trap.
The categorical distribution is the special case of the multinomial with N = 1. 

Poisson - Used in SCR as a prior on population size N (Poisson point process of activity centers), model for frequency of captures in a trap (for when an individual can be detected multiple times at a trap in a single occassion).

Uniform distribution - often used as a prior for detection probability or to set a uniform prior for the coordinates of points in space. 

Bivariate Normal - used as a rudimentary description of movement outcomes about a home range center.

SCR - any individual encounter model where spatial information is also obtained.
SCR models have the random variables $s$, $u$, and $y$
$s$ is the activity center
$u$ is the location of the individual at time of sampling
$y$ is the encounter data
$n(y)$ refers to a summary of y, which is sometimes all that is observed
Thus, SCR can be succintly described as:
$$[n(y)|y][y|u][u|s][s]$$

Lets break down each submodel

$[s]$ is a point process of the activity centers

$[u|s]$ is the location of an individual given their activity center, this sub model can be ignored if movement is not of particular interest or s is defined in a way not related to home range centers

$[y|s]$ describes how capture histories arise conditional on the locations of individuals, when there is no movement model the submodel is instead $[y|s]$ . This model has a detection probability that varies with distance from $s$ as a function of two parameters $p0$ and $\sigma$, the detection probability at the activity center, and the decay of the probability over space.

Simplest SCR Model
$$ N \sim Poisson(\mu||S||) $$
$$ s_i \sim Uniform(S) $$
$$ y_{ijk} |s_i \sim Bernoulli(p(||x_j - s_i ||))$$ 

## Chapter 3
Covers GLMMs in a Bayesian framework, primarily in the BUGS language.
This material has also been covered in AHM and FW599. 
GLMMS - linear model with a link function and a distribution from the exponential family. 

Bayes Rule:
$$ [\theta|y] = \frac{[y|\theta][\theta] }{[y]} $$

When settng the prior $N$ for the number of individuals in the population
$$ Binomial(M, \psi) $$
the prior $N$ needs to be sufficiently large to not bound the estimation of $M$

REMEMBER: priors are not invariant to link function transformations!
Bayesian inference is valid for all sample sizes, not only asymptotically valid at large enough sample sizes like frequentist techniques.

Review of gibbs sampling, Metropolis-Hastings, BUGS Language syntax, convergence diagnositics, burn in, effective sample size, CrI.

Monte-Carlo error should be small relative to the magnitude of the parameter estimate. 

Poisson GLMS: Basic Poisson GLM, the canonical model for counts in ecology, with a fixed mean-variance relationship that is often problematic.

```{r, eval = F}

model{
  for(i in 1:nObs){
    y[i] ~dpois(lambda[i])
    log(lambda[i]) <- beta0 + beta1*variable1[i]
  }
  beta0 ~ dunif(-5,5)
  beta1 ~ dunif(-5,5)
}

```

Poisson GLMM with OLRE (Log-Normal Mixture)
Used to deal with extra poisson variation (overdispersion)
```{r, eval = F}
model{
  #Likelihood
  for(i in 1:nObs){
    y[i] ~dpois(lambda[i])
    log(lambda[i]) <- beta0 + beta1*variable1[i] + eta[i]
    eta[i] ~dnorm(0, tau)
  }

  #Priors
  beta0 ~ dunif(-5,5)
  beta1 ~ dunif(-5,5)
  tau ~ 1/(sigma*sigma)
  sigma ~ dunif(0,10)
}
```

Binomial GLMs 

When you have bounded counts or a single trial with success probability p (Bernoulli case).
N-mixture models are a special case where the number of trials itself is a latent parameter. 
In capture recapture modeling the number of individuals captured out of the total population size is a binomial estimation problem.
Logit and cloglog are typical link functions, cloglog is useful when the response should scale with area or effort. 

Complementary Log Log link
$$ cloglog(p_i) = log(-log(l-p_i)) = log(\lambda_i) $$

This naturally connects detection probability with local abundance, which is a logical connection. 

Bayesian Model Checking - bayesian p value is authors choice, "create a discrepancy measure of the observed and predicted data, then compare the distribution of that measure to the piosterior predictive distribution of that statistics for hypothetical perfect datasets where the model is correct."

Pearson residuals descrepancy statistic for count frequency data
$$ D(y_i, \theta) = \frac{y_i -\mathbb{E}(y_i)}{\sqrt{Var(y_i)}} $$

Fit Statistic is generated from the squared residuals computed from the observations
$$ T(\textbf{y}, \theta) = \sum_i D(y_i, \theta)^2 $$

This fit statistic is computed at every iteration of a MCMC algorithm, given current parameter values. At each iteration, you also simulate a new dataset given the current parameters of the MCMC, then generate the same fit statistic

$$ T(\textbf{y}^{new}, \theta) = \sum_i D(y_i^{new}, \theta)^2 $$

Then, the bayesian p-value is 
$$ Pr(T(y^{new})) > T(y) $$

which should be close to 0.5

Model Selection - three methods
First appraoch - use the probability of direction of a variable to assess the strength of it for inclusion, can be problematic for some levels of inferential rigor, but is very useful and simple for tasks of lower causal strength such as prediction.

Indicator variable approach :
$$ \mathbb{E}(y_i) = \beta_0 + w_1\beta_1x_i $$
$$ w_1 \sim Bernoulli(0.50) $$

The posterior density of $w_1$ = 1 or = 0 gives the relative support for inclusion in the model. 
If there are diffuse priors for the structural variables of the model, the chain may very infrequently jump from $w_1$ = 1 to = 0 infrequently, leading to poor mxing. 
A solution is to fit the full model without weight variables, then use those posteriors as priors for the structural variables in a model selection run. 
The third approach is to use subject-matter expertise and context to select models that are biologically and structurally plausible. 

## Chapter 4
Closed Capture-Recapture Population Models

Needs to Be Updated - ALN Read this chapter on a mobile device on a plane and was unable to take notes

## Chapter 5
Spatial Capture Recapture Models

The prior chapter introduced a primitive scr model, where the distance from home range center to trapping array boundary was an individual covariate. 
For full SCR, the location of the animal and each individual detector is included.
Instead of distance to array edge, we define the distance to each individual trap.
Input Data is JxKxN array of j traps, across K occassions, for each N indiviudal.

Model SCR0 - most applicable to passive devices like hare snares, assumes: devices don't fill up, individual can be captured by more than one trap in an occasion, and individuals are detected at most 1 time per detector per occasion.

The simplest version, with no covariates, collapses occasions into a count of detections out of K sampling sessions, and thus detection histories of individuals at each trap are
$$ y_{ij} \sim Binomial(K, p_{ij}) $$
Note here that we are indexing by $i$ individual and $j$ trap, as opposed to a cr model that would just index by individual.

$p_{ij}$ varies based on the distance between the trap and the home range center of that indvidual.
Distance can be related to detection probability with a logistic glm, or with the half normal detection function like in distance sampling, except $p_0$ can be less than 1 in SCR, unlike in distance sampling. 
For different detection functions, their behavior as they approach zero varies, but they generally are all monotonically decreasing with distance in a non-linear fasion. 
Figure 5.1 effectively shows the differences between gaussian, gaussian hazard, negative exponential, and logistic.
In SCR, home range is just defined as the area used by an organism during an arbitrary time period, and the center is just the centroid, which could be asymptotically observed with many VHF fixes.
The distance of each observation from the individual's home range center is unobserved, and thus requires a prior distribution, the uniform assumption of home range centers in space leads to a uniform prior on s.
$s$ is a binomial point process, since the number of total points is fixed unlike a poisson point process.
The total number of points, $N$ is fixed because we add $M$ fake individuals in our data augmentation procedure and thus know the upper bound of possible individuals.
Realizations of this point process will result in patterns that are not uniform, but do exhibit some spatial clustering, because even though the priors are uniform, we are informing them with data which leads to a non-uniform posterior.

Note that the state space in SCR is often called an observation window in the point process literature. $S$ is the state space of $s_i$. 
$$ s_i \sim Uniform(S) $$
 
Choosing $S$ can seem arbitrary, but if it is large enough to include all individuals, then density is invariant to $S$, as it scales with $N$ given constant density. 
Increasing $S$ increases computational complexity because $M$ total individuals including augmented individuals must increase. 
You can also think in terms of distance sampling, where if detection decreases with distance, you could create an arbitrary number of sampling distance boxes way far out and they would contribute nothing as the expected number of detected individuals will essentially be 0.
Creating irregular state spaces can be difficult in bugs, but is easier in secr. 

The detection process is a rudimentary model of animal movement, and detection can be broken down into detection given the animal is using the area conditional on the probability of usage.
Thus, if using a negative exponential detection function, the detection function is identical to a RSF. 
Therefore, if we parameterize detection with habitat covariates, we are integrating an RSF into our SCR model, although I am curious if the way the show integrating an RSF will be more formal than this.
Telemetry can be integrated into a SCR model here with the detection function.
Under the bivariate normal detection model we can compute home range area directly, by using sigma.
This is particularly useful in reverse when conducting simulation studies, because you can calculate what sigma should be to result in a given home range area that can be motivated by prior knowledge.
Chapter 12 covers least cost path models of distance, which will likely be useful for voles as they are traveling along trails.
The authors provide useful code for simulating SCR data and the simSCR0 function in the scrbook package can be used.

Basic SCR0 model code:

```{r, eval = F}
cat("
model{

  #Priors
  alpha0 ~ dnorm(0,.1) # alpha prior
  logit(p0) <- alpha0 #baseline detection probability prior
  alpha1 ~ dnorm(0,.1) #distance decay prior
  sigma <- sqrt(1/(2*alpha1)) # sigma conversion
  psi ~ dunif(0,1) # Data Augmentation proportion prior

  #Likelihood
  for(i in 1:M){

    z[i] ~ dbern(psi) #Data augmentation parameter

    s[i,1] ~ dunif(xlim[1],xlim[2]) # Home range X coord
    s[i,2] ~ dunif(ylim[1],ylim[2]) # Home range y coord

      for(j in 1:J){

        d[i,j] <- pow(pow(s[i,1]-X[j,1],2) + pow(s[i,2]-X[j,2],2),0.5) #distance of this encounter
        p[i,j] <- z[i]*p0*exp(- alpha1*d[i,j]*d[i,j]) #probability of detection
        y[i,j] ~ dbin(p[i,j],K) # observed detection history
        
      }
  }

  # Derived Parameters
  N <- sum(z[]) # number of indviduals
  D <- N/64 # Density
}
",file = "SCR0b.txt")
```

The implied home range area can vary a decent bit based on what detection function we choose, but N and density are less sensitive to this choice.

Estimating N or density can be broken into three objectives, the value for N for the population that was sampled, also known as realized N, N for a similar but unsampled population, and the expected value of N.

SCR0 Model Assumptions: demographic and geographic closure, activity centers are randomly distributed, detection is a function of distance, encounters among and of the same indivdiual are independent.

Wolverine worked example: camera trapping study in SE Alaska
Data consist of an ecnounter data file and trap deployment file, which manage which individuals were seen at which traps and which traps were operating for what days and at what locations.
This data structure will be relatively standard across SCR studies.

When defining a state space, it is useful to standardize the coordinates and have it defined near the origin so the chains mix well.
In the wolverine example, the state space is defined as a continuous space within a rectangle, but do deal with masking it is easier to discretize the space into points and mask out points that are unable to contain individuals, such as the ocean for wolverines.
Note, when discretizing space be cautious, a home range center in the middle of a parking lot could be a good estimate if there is great habitat around the parking lot, and a small mammal travels around the perimeter of that lot to forage.
In the SCR0 model above, the state space is defined by the uniform priors on the x and y coordinates of each activity center.
In this wolverine examople, and in other cases, if an individudal is moving from detector to detector to harvest bait, it can artificially alter the movement behavior of the individual and thus the resulting home range size estimate.
When generating a discretized state space, there is a trade off in the resolution of the grid of points and the computational intensity.
Density maps, home range locations, etc. are all derived parameters and can be calculated relatively easily from an SCR model.
Note, when creating a map of density, it will tend toward the mean density as you get far away from detectors, as the influence of the data weakens.

 
## Chapter 6
Likelihood analysis of SCR

Introduces likelihood fitting of SCR models, primarily in secr. 
Likelihood fitting relies on the integrated likelihood, which is based on the marginal distribution of the data, y, in which the random effects
are removed by integration from the conditional-on-s distribution of the observations. 
These are removed by integrating across the state space $S$. 
In the chapter example, the integral is calculated by the rectangular rule, with an equal area mesh of points over the state space. 
Essentially, we are evaluating the probability of the data given all possible values of $s$, the activity center location for that individual. 
Note when comparing a MLE and Bayesian estimate of a parameter, the Bayesian multivariate posterior mode will be the MLE estimate assuming flat priors, but since the mean of the posterior is often reported, there will be some discrepancy. 
The authors first demonstrate an example where N is known, then when N is unknown. 
To deal with unknown N, you introduce a comninatoric term that finds the number of $n_0$ for some unknown $N$. 
Another approach is data augmentation, but authors suggest that it is better suited for Bayesian analyses.
It is more difficult to fit models with covariates on parameters such as detection probability if the values for the covariate are not known for every individual in a likelihood based approach. 
It is rather trivial for a Bayesian approach as values are imputed from the prior. 
Authors suggest to avoid model selection on different distance functions for detection, as there are rarely good reasons to choose one over the other, and model selection is likely to lead to overfitting in these cases. 
In general, goodness of fit tests in SCR models with realistic sample sizes may be underpowered and unable to detect departures from the prescribed model. 
Uncertainty in derived parameters has to be calculated with the delta method for likelihood based approaches. 
To add a habitat mask, the integration grid can be explicitly created to only occur in areas of potential habitat.

secr is the Efford package for likelihood estimation of scr models.
The basic data structure is a detectors object and a captures object.
Detector types include proximity (hair snare), count (camera trap), multicapture (pitfall), single (longworth trap).
Multi-sesion models are when there are multiple studied populations, or some other stratification, and can be handled in secr
For single catch traps, the encounter model is a multinomial as opposed to the bernoulli encounter model. 
SECR can provide helpful summaries of capture data such as mean-maximum distance moved, will provide automatic warnings if the state space is too small, will do model selection with AIC, will implement a population closure test, etc.

## Extension: An integrated path for spatial capture–recapture and animal movement modeling
McClintock et al. 2021 Ecology  https://doi.org/10.1002/ecy.3473

This article introduces and special issue on movement modeling and SCR.
Typically, SCR focuses on population level processes, while movement modeling focuses on individual level processes.

Terminology: Langragian - individual, Eulerian - population, level processes

Integrating realistic movement processes can reduce bias in population parameter estimation from SCR. 
The key methodological approach is incorporating a mvoement model into SCR so that observations are conditioned on the dynamic location of an individual as opposed to a static and latent activity center.

Basics of SCR <br>
State Space $M$ is the area in 2 dimensional space where an individual resides during the study.
Sampling occurs by individually identifying individuals detected at an array of $J$ detectors over $T$ time intervals, although other sampling designs are possible such as search encounter and continuous time.
The spatial encounter history of each individual is the key data stream collected.
$$ [y|s,N,\theta] = [y|s,\theta] [s, N|\theta] $$

The first term is the $y$ encounters, conditional on $s$, activity centers, $N$ the number of individuals in the state space, $\theta$ and fitted parameters. 
The second term in the SCR model itself where the locations of individuals $y$ is conditional on their activity centers $s$ and how quickly detection decays $\theta$. 
The activity centers are described by the point process model, where the activity centers for N individuals, $s,N$, are conditional on the point process parameters $\theta$.
The detection of an individual is a function of the difference in location between the detector and the latent activity center:

$$ p_{ij} = p_0 exp(-\frac{\parallel x_j - s_i \parallel^2}{2 \sigma^2_{det}}) $$

Here $p0$ is the baseline detection probability and $\sigma$ controls the decay rate, which is symmetric and gaussian. 
The $\parallel^2%$ here refers to a bivariate normal distribution, not the distance squared.

Under the point process model, a homogenous poisson point process assumes that activity centers are uniformly distributed across space, while using an inhomogenous point process allows for the intensity (number of indviduals) to vary with some underlying covariate like habitat condition. 
Figure 2 provides an excellent overview of the process and data model of SCR and a movement integrated SCR. 

Extensions include allowing for open populations (Schaub and Royle 2014), resource selection, landscape connectivity, territoriality, attraction, ipms, passive acoustics, and unmarked individuals. For these extensions, they are implemented as either inhomogenous point process models, or as covariates on $\sigma$ and $p0$. Another rudimentary extension to include movement is to allow activity centers that change based on a simple random walk between primary periods. 

Movement Modeling<br>
Common movement modeling techniques include RSFs and SSFs, and those can be extended to spatiotemporal point process models to distinguish resourece selection from availability while accounting for the underlying process of movement (such as different movement speeds in different habitats).
Lagrangian (individual) movement models are parameterized random walk models in discrete or continuous time. 
Simple discrete time movement model
$$\mu_t = \mu_{t-1} + \epsilon_t $$
where $\epsilon$ is a symmetric normal dispersion kernel.
This generates a time series of points that walks around, see Fig 3

Additional extensions to the random walk can be easily incorporated, such as bias to a central place:
$$ \mu_t = \mu_{t-1} + \boldsymbol B(a-\mu_{t-1}) + \epsilon_t $$
where $\boldsymbol B$ is a matrix that governs the degree of attraction to point a.

Adding correlation in direction between the steps creates longer, more realistic move similar to a non central place forager searching through a landscape. 
More complex and modern movement models include multiple states, each with their own set of movement parameters, which can be easily done with a piecewise equation for $n$ states.
Movement parameters can be a function of covariates such as age and sex. 

Potential functions allow for the modeling of movement along environmental gradients, the gradient is a vector field of partial derivatives that convey the direction and magnitude of attraction. 

Limiting Distributions<br>
When going from Lagrangian vs Eulerian, the limiting distribution of the time series process is a useful concept, which is the limit of $$ \mu_{t+h} | \mu_t $$ as h -> $\infty$

Conceptually, this is creating greater and greater time gaps between locations, so that they are independent. This limiting distribution of $N$ individuals in $M$ state space is an animal density surface, and is thus conceptually united to SCR model parameters. 

Few movement models have closed form limiting distributions, or limiting distributions at all, but the biased random walk does, see Eqn 11. 

The SCR observation model can be a non-uniform thinning of the limiting distribution of the movement of all individuals in $M$, this is most realistic when an animal can move throughout its entire home range within an occasion. 
The Langevin diffusion RSF model also has a closed form limiting distribution. 

Movement data has GPS error, and is thus often modeled hierarchically with a movement and measurement sub model. 

The scale parameter assumptions have led SCR to be most widely used with species with symmetric and resident home ranges, as opposed to nomadic species. 
Uncertainty can be underestimated, even though abundance is not biased when the modeled species has a non-circular home range. 
Considerable bias can occur with sigma (think dune example from Taal) as observations that are very far away assume a large circle of higher detection probability. 
Resource selection can bias density becuase detection is not just a single function of distance. 
Movement also becomes extremely important in open population model parameter estimation. 
Animal movement models are not typically attached to demographic rates or population estimates. 

Integrating SCR and Movement Data<br>
Basic integrated SCR model can be described as
$$[y|\mu, N, \theta] = [y|\mu, \theta ] [\mu | \theta] [\mu_1, N, |\theta] $$
Term 1 is SCR. 
Term 2 is the movement model.
Term 3 is the point process model.
$\mu$ the position process, is now a latent variable, as we do not know the exact position of the animal in each time step. 
Thus, we estimate a latent movement trajectory as opposed to a static activity center.
Therefore, the distance for the detection kernel becomes the distance between the detector and the latent location at time $t$ as opposed to the activity center $s$. 
The point process now represents the initial positon of N individuals as opposed to N activity centers.
By adding a center of attraction, one can estimate second and third order resource selection (selection of a home range within $M$ and selection of resources within a home range).

One issue with uniting movement data and SCR data is a discrete time versus continuous time mismatch.
One solution is to model the distance between the expected location of the individual within the time step and the observed location, this seems conceptually similar to a dynamic activity center (it is currently unclear to me how it is not the same thing).

Next Steps <br>
One challenge with movement-scr integrated models is optimal sampling is unknown. 
In SCR, the sampling trade of is between the number of individuals captured and the extent of spatial recaptures, which is often dictated by trap spacing and is typically suggested to be $1-2\sigma$. 
Without auxiliary location data, the number of spatial recaptures required is likely to be higher in order to more appropriately model the movement process.

Another challenge is that sampling bias is controlled in SCR through the detection process, but sampling bias is typically not corrected for in movement studies, with telemetered individuals not representing a random sample of the population, and thus potentially biasing the parameters of the movement model from the true population mean.

Models can be broken down into 4 quadrants, space and time bring either discrete or continous. 

Movement near the boundaries of the state space can be a challenge, as simple random walk difffusive movement would lead to a deterministic reduction in density over time that is unrealistic.
A simple solution is expanding the state space far beyond the boundaries of the movement of individuals, which can lead to longer model fitting processes due to the need to estimate more parameters. 
Additionally, now that each movement trajectory is latent as opposed to just the home range center, there are $NT$ latent parameters to estimate as opposed to simply $N$.  
