---
title: "Austin Nash's Notes for Spatial Capture Recapture"
format:
  html:
    embed-resources: true
theme: darkly
---

## Chapter 1
SCR is an extension to capture recapture methods, but also allows for the integration of a wide array of data types and ecological theory.
Common data streams used in SCR analyses include camera traps, DNA sampling, and acoustic recorders.
While typical data streams revolve around static detectors, search encounter methods such as detection dog surveys can be used in an SCR framework.
Encounter histories are commonly represented by a matrix with rows representing individuals and columns representing sampling events.
Authors present a motivating example of using traditional CR methods to estimate the density of bears. 
Going from N individuals to density requires strong assumptions, and density estimates vary widely based on the method selected. 
Traditional methods can't account for differences in exposure to detectors of individuals based on the location of their activity centers.
Traditional CR cannot accomadate trap level covariates. 
THe classical approach to estimating density involves adding a buffer to the MCP of the trap grid that is 0.5 times the size of the mean maximum distance moved. 
Another approach to correcting biases in traditional CR models is by using telemetry to estiamte how often an animal temporarily emigrates from the trapping area. 
Another CR bias correction approach is using distance to edge of the trapping grid as an individual covariate on detectability, which is an intermediate step to full SCR.

Efford 2004 is the first formulation of a spatially explicit CR. 
This involved a point process model formulation and equated the observation process to a distance sampling problem. 
Observations of individuals represent a imperfectly observed, or thinned, point process. 
Another formalization is activity centers being parent nodes, with locations of individuals being child nodes, which is a poisson cluster process. 
Authors use the term home range and activity center interchangeably. 
SCR does not require the organism to be a central place forager, the activity center is essentially equivalent to a utilization distribution. 
The possible values of all activity center points are referred to as the state space. 
$$D = \frac {N}{A(S)} $$

## Chapter 2
Primarily review, covers the basics of statistical inference.
Review of probability notation, random variables, common distributions

Binomial Distribution - used in SCR as an encounter model for the number of times i individual is captured at trap j after k occasions. Probability of encounter is governed by the distance between the activity center and the detector. Also used as a prior on population size. 

Bernoulli Distribution - Binomial where n = 1, proper distribution to use when individuals have different detection probabilities (in the distance case have different detection kernels)

Multinomial distribution - commonly used in CR studies to model the encounter histories of individuals. 
Each category is each possible encounter history. 
The multinomial is used in scr to model when an individual can only be caught in a single trap in an ocassion (like for box traps, unlike camera traps), N = 1 and the cell probabillities are the probability of being captured in each trap.
The categorical distribution is the special case of the multinomial with N = 1. 

Poisson - Used in SCR as a prior on population size N (Poisson point process of activity centers), model for frequency of captures in a trap (for when an individual can be detected multiple times at a trap in a single occassion).

Uniform distribution - often used as a prior for detection probability or to set a uniform prior for the coordinates of points in space. 

Bivariate Normal - used as a rudimentary description of movement outcomes about a home range center.

SCR - any individual encounter model where spatial information is also obtained.
SCR models have the random variables $s$, $u$, and $y$
$s$ is the activity center
$u$ is the location of the individual at time of sampling
$y$ is the encounter data
$n(y)$ refers to a summary of y, which is sometimes all that is observed
Thus, SCR can be succintly described as:
$$[n(y)|y][y|u][u|s][s]$$

Lets break down each submodel

$[s]$ is a point process of the activity centers

$[u|s]$ is the location of an individual given their activity center, this sub model can be ignored if movement is not of particular interest or s is defined in a way not related to home range centers

$[y|s]$ describes how capture histories arise conditional on the locations of individuals, when there is no movement model the submodel is instead $[y|s]$ . This model has a detection probability that varies with distance from $s$ as a function of two parameters $p0$ and $\sigma$, the detection probability at the activity center, and the decay of the probability over space.

Simplest SCR Model
$$ N \sim Poisson(\mu||S||) $$
$$ s_i \sim Uniform(S) $$
$$ y_{ijk} |s_i \sim Bernoulli(p(||x_j - s_i ||))$$ 

## Chapter 3
Covers GLMMs in a Bayesian framework, primarily in the BUGS language.
This material has also been covered in AHM and FW599. 
GLMMS - linear model with a link function and a distribution from the exponential family. 

Bayes Rule:
$$ [\theta|y] = \frac{[y|\theta][\theta] }{[y]} $$

When settng the prior $N$ for the number of individuals in the population
$$ Binomial(M, \psi) $$
the prior $N$ needs to be sufficiently large to not bound the estimation of $M$

REMEMBER: priors are not invariant to link function transformations!
Bayesian inference is valid for all sample sizes, not only asymptotically valid at large enough sample sizes like frequentist techniques.

Review of gibbs sampling, Metropolis-Hastings, BUGS Language syntax, convergence diagnositics, burn in, effective sample size, CrI.

Monte-Carlo error should be small relative to the magnitude of the parameter estimate. 

Poisson GLMS: Basic Poisson GLM, the canonical model for counts in ecology, with a fixed mean-variance relationship that is often problematic.

```{r, eval = F}

model{
  for(i in 1:nObs){
    y[i] ~dpois(lambda[i])
    log(lambda[i]) <- beta0 + beta1*variable1[i]
  }
  beta0 ~ dunif(-5,5)
  beta1 ~ dunif(-5,5)
}

```

Poisson GLMM with OLRE (Log-Normal Mixture)
Used to deal with extra poisson variation (overdispersion)
```{r, eval = F}
model{
  #Likelihood
  for(i in 1:nObs){
    y[i] ~dpois(lambda[i])
    log(lambda[i]) <- beta0 + beta1*variable1[i] + eta[i]
    eta[i] ~dnorm(0, tau)
  }

  #Priors
  beta0 ~ dunif(-5,5)
  beta1 ~ dunif(-5,5)
  tau ~ 1/(sigma*sigma)
  sigma ~ dunif(0,10)
}
```

Binomial GLMs 

When you have bounded counts or a single trial with success probability p (Bernoulli case).
N-mixture models are a special case where the number of trials itself is a latent parameter. 
In capture recapture modeling the number of individuals captured out of the total population size is a binomial estimation problem.
Logit and cloglog are typical link functions, cloglog is useful when the response should scale with area or effort. 

Complementary Log Log link
$$ cloglog(p_i) = log(-log(l-p_i)) = log(\lambda_i) $$

This naturally connects detection probability with local abundance, which is a logical connection. 

Bayesian Model Checking - bayesian p value is authors choice, "create a discrepancy measure of the observed and predicted data, then compare the distribution of that measure to the piosterior predictive distribution of that statistics for hypothetical perfect datasets where the model is correct."

Pearson residuals descrepancy statistic for count frequency data
$$ D(y_i, \theta) = \frac{y_i -\mathbb{E}(y_i)}{\sqrt{Var(y_i)}} $$

Fit Statistic is generated from the squared residuals computed from the observations
$$ T(\textbf{y}, \theta) = \sum_i D(y_i, \theta)^2 $$

This fit statistic is computed at every iteration of a MCMC algorithm, given current parameter values. At each iteration, you also simulate a new dataset given the current parameters of the MCMC, then generate the same fit statistic

$$ T(\textbf{y}^{new}, \theta) = \sum_i D(y_i^{new}, \theta)^2 $$

Then, the bayesian p-value is 
$$ Pr(T(y^{new})) > T(y) $$

which should be close to 0.5

Model Selection - three methods
First appraoch - use the probability of direction of a variable to assess the strength of it for inclusion, can be problematic for some levels of inferential rigor, but is very useful and simple for tasks of lower causal strength such as prediction. 
Indicator variable approach :
$$ \mathbb{E}(y_i) = \beta_0 + w_1\beta_1x_i $$
$$ w_1 \sim Bernoulli(0.50) $$