---
title: "Austin Nash's Notes for Spatial Capture Recapture"
format:
  html:
    embed-resources: true
theme: darkly
---

## Chapter 1
SCR is an extension to capture recapture methods, but also allows for the integration of a wide array of data types and ecological theory.
Common data streams used in SCR analyses include camera traps, DNA sampling, and acoustic recorders.
While typical data streams revolve around static detectors, search encounter methods such as detection dog surveys can be used in an SCR framework.
Encounter histories are commonly represented by a matrix with rows representing individuals and columns representing sampling events.
Authors present a motivating example of using traditional CR methods to estimate the density of bears. 
Going from N individuals to density requires strong assumptions, and density estimates vary widely based on the method selected. 
Traditional methods can't account for differences in exposure to detectors of individuals based on the location of their activity centers.
Traditional CR cannot accomadate trap level covariates. 
THe classical approach to estimating density involves adding a buffer to the MCP of the trap grid that is 0.5 times the size of the mean maximum distance moved. 
Another approach to correcting biases in traditional CR models is by using telemetry to estiamte how often an animal temporarily emigrates from the trapping area. 
Another CR bias correction approach is using distance to edge of the trapping grid as an individual covariate on detectability, which is an intermediate step to full SCR.

Efford 2004 is the first formulation of a spatially explicit CR. 
This involved a point process model formulation and equated the observation process to a distance sampling problem. 
Observations of individuals represent a imperfectly observed, or thinned, point process. 
Another formalization is activity centers being parent nodes, with locations of individuals being child nodes, which is a poisson cluster process. 
Authors use the term home range and activity center interchangeably. 
SCR does not require the organism to be a central place forager, the activity center is essentially equivalent to a utilization distribution. 
The possible values of all activity center points are referred to as the state space. 
$$D = \frac {N}{A(S)} $$

## Chapter 2
Primarily review, covers the basics of statistical inference.
Review of probability notation, random variables, common distributions

Binomial Distribution - used in SCR as an encounter model for the number of times i individual is captured at trap j after k occasions. Probability of encounter is governed by the distance between the activity center and the detector. Also used as a prior on population size. 

Bernoulli Distribution - Binomial where n = 1, proper distribution to use when individuals have different detection probabilities (in the distance case have different detection kernels)

Multinomial distribution - commonly used in CR studies to model the encounter histories of individuals. 
Each category is each possible encounter history. 
The multinomial is used in scr to model when an individual can only be caught in a single trap in an ocassion (like for box traps, unlike camera traps), N = 1 and the cell probabillities are the probability of being captured in each trap.
The categorical distribution is the special case of the multinomial with N = 1. 

Poisson - Used in SCR as a prior on population size N (Poisson point process of activity centers), model for frequency of captures in a trap (for when an individual can be detected multiple times at a trap in a single occassion).

Uniform distribution - often used as a prior for detection probability or to set a uniform prior for the coordinates of points in space. 

Bivariate Normal - used as a rudimentary description of movement outcomes about a home range center.

SCR - any individual encounter model where spatial information is also obtained.
SCR models have the random variables $s$, $u$, and $y$
$s$ is the activity center
$u$ is the location of the individual at time of sampling
$y$ is the encounter data
$n(y)$ refers to a summary of y, which is sometimes all that is observed
Thus, SCR can be succintly described as:
$$[n(y)|y][y|u][u|s][s]$$

Lets break down each submodel

$[s]$ is a point process of the activity centers

$[u|s]$ is the location of an individual given their activity center, this sub model can be ignored if movement is not of particular interest or s is defined in a way not related to home range centers

$[y|s]$ describes how capture histories arise conditional on the locations of individuals, when there is no movement model the submodel is instead $[y|s]$ . This model has a detection probability that varies with distance from $s$ as a function of two parameters $p0$ and $\sigma$, the detection probability at the activity center, and the decay of the probability over space.

Simplest SCR Model
$$ N \sim Poisson(\mu||S||) $$
$$ s_i \sim Uniform(S) $$
$$ y_{ijk} |s_i \sim Bernoulli(p(||x_j - s_i ||))$$ 

## Chapter 3
Covers GLMMs in a Bayesian framework, primarily in the BUGS language.
This material has also been covered in AHM and FW599. 
GLMMS - linear model with a link function and a distribution from the exponential family. 

Bayes Rule:
$$ [\theta|y] = \frac{[y|\theta][\theta] }{[y]} $$

When settng the prior $N$ for the number of individuals in the population
$$ Binomial(M, \psi) $$
the prior $N$ needs to be sufficiently large to not bound the estimation of $M$

REMEMBER: priors are not invariant to link function transformations!
Bayesian inference is valid for all sample sizes, not only asymptotically valid at large enough sample sizes like frequentist techniques.

Review of gibbs sampling, Metropolis-Hastings, BUGS Language syntax, convergence diagnositics, burn in, effective sample size, CrI.

Monte-Carlo error should be small relative to the magnitude of the parameter estimate. 

Poisson GLMS: Basic Poisson GLM, the canonical model for counts in ecology, with a fixed mean-variance relationship that is often problematic.

```{r, eval = F}

model{
  for(i in 1:nObs){
    y[i] ~dpois(lambda[i])
    log(lambda[i]) <- beta0 + beta1*variable1[i]
  }
  beta0 ~ dunif(-5,5)
  beta1 ~ dunif(-5,5)
}

```

Poisson GLMM with OLRE (Log-Normal Mixture)
Used to deal with extra poisson variation (overdispersion)
```{r, eval = F}
model{
  #Likelihood
  for(i in 1:nObs){
    y[i] ~dpois(lambda[i])
    log(lambda[i]) <- beta0 + beta1*variable1[i] + eta[i]
    eta[i] ~dnorm(0, tau)
  }

  #Priors
  beta0 ~ dunif(-5,5)
  beta1 ~ dunif(-5,5)
  tau ~ 1/(sigma*sigma)
  sigma ~ dunif(0,10)
}
```

Binomial GLMs 

When you have bounded counts or a single trial with success probability p (Bernoulli case).
N-mixture models are a special case where the number of trials itself is a latent parameter. 
In capture recapture modeling the number of individuals captured out of the total population size is a binomial estimation problem.
Logit and cloglog are typical link functions, cloglog is useful when the response should scale with area or effort. 

Complementary Log Log link
$$ cloglog(p_i) = log(-log(l-p_i)) = log(\lambda_i) $$

This naturally connects detection probability with local abundance, which is a logical connection. 

Bayesian Model Checking - bayesian p value is authors choice, "create a discrepancy measure of the observed and predicted data, then compare the distribution of that measure to the piosterior predictive distribution of that statistics for hypothetical perfect datasets where the model is correct."

Pearson residuals descrepancy statistic for count frequency data
$$ D(y_i, \theta) = \frac{y_i -\mathbb{E}(y_i)}{\sqrt{Var(y_i)}} $$

Fit Statistic is generated from the squared residuals computed from the observations
$$ T(\textbf{y}, \theta) = \sum_i D(y_i, \theta)^2 $$

This fit statistic is computed at every iteration of a MCMC algorithm, given current parameter values. At each iteration, you also simulate a new dataset given the current parameters of the MCMC, then generate the same fit statistic

$$ T(\textbf{y}^{new}, \theta) = \sum_i D(y_i^{new}, \theta)^2 $$

Then, the bayesian p-value is 
$$ Pr(T(y^{new})) > T(y) $$

which should be close to 0.5

Model Selection - three methods
First appraoch - use the probability of direction of a variable to assess the strength of it for inclusion, can be problematic for some levels of inferential rigor, but is very useful and simple for tasks of lower causal strength such as prediction.

Indicator variable approach :
$$ \mathbb{E}(y_i) = \beta_0 + w_1\beta_1x_i $$
$$ w_1 \sim Bernoulli(0.50) $$

The posterior density of $w_1$ = 1 or = 0 gives the relative support for inclusion in the model. 
If there are diffuse priors for the structural variables of the model, the chain may very infrequently jump from $w_1$ = 1 to = 0 infrequently, leading to poor mxing. 
A solution is to fit the full model without weight variables, then use those posteriors as priors for the structural variables in a model selection run. 
The third approach is to use subject-matter expertise and context to select models that are biologically and structurally plausible. 

## Chapter 4
Closed Population Models

## Extension An integrated path for spatial capture–recapture and animal movement modeling
McClintock et al. 2021 Ecology  https://doi.org/10.1002/ecy.3473

This article introduces and special issue on movement modeling and SCR.
Typically, SCR focuses on population level processes, while movement modeling focuses on individual level processes.

Terminology: Langragian - individual, Eulerian - population, level processes

Integrating realistic movement processes can reduce bias in population parameter estimation from SCR. 
The key methodological approach is incorporating a mvoement model into SCR so that observations are conditioned on the dynamic location of an individual as opposed to a static and latent activity center.

Basics of SCR <br>
State Space $M$ is the area in 2 dimensional space where an individual resides during the study.
Sampling occurs by individually identifying individuals detected at an array of $J$ detectors over $T$ time intervals, although other sampling designs are possible such as search encounter and continuous time.
The spatial encounter history of each individual is the key data stream collected.
$$ [y|s,N,\theta] = [y|s,\theta] [s, N|\theta] $$

The first term is the $y$ encounters, conditional on $s$, activity centers, $N$ the number of individuals in the state space, $\theta$ and fitted parameters. 
The second term in the SCR model itself where the locations of individuals $y$ is conditional on their activity centers $s$ and how quickly detection decays $\theta$. 
The activity centers are described by the point process model, where the activity centers for N individuals, $s,N$, are conditional on the point process parameters $\theta$.
The detection of an individual is a function of the difference in location between the detector and the latent activity center:

$$ p_{ij} = p_0 exp(-\frac{\parallel x_j - s_i \parallel^2}{2 \sigma^2_{det}}) $$

Here $p0$ is the baseline detection probability and $\sigma$ controls the decay rate, which is symmetric and gaussian. 
The $\parallel^2%$ here refers to a bivariate normal distribution, not the distance squared.

Under the point process model, a homogenous poisson point process assumes that activity centers are uniformly distributed across space, while using an inhomogenous point process allows for the intensity (number of indviduals) to vary with some underlying covariate like habitat condition. 
Figure 2 provides an excellent overview of the process and data model of SCR and a movement integrated SCR. 

Extensions include allowing for open populations (Schaub and Royle 2014), resource selection, landscape connectivity, territoriality, attraction, ipms, passive acoustics, and unmarked individuals. For these extensions, they are implemented as either inhomogenous point process models, or as covariates on $\sigma$ and $p0$. Another rudimentary extension to include movement is to allow activity centers that change based on a simple random walk between primary periods. 

Movement Modeling<br>
Common movement modeling techniques include RSFs and SSFs, and those can be extended to spatiotemporal point process models to distinguish resourece selection from availability while accounting for the underlying process of movement (such as different movement speeds in different habitats).
Lagrangian (individual) movement models are parameterized random walk models in discrete or continuous time. 
Simple discrete time movement model
$$\mu_t = \mu_{t-1} + \epsilon_t $$
where $\epsilon$ is a symmetric normal dispersion kernel.
This generates a time series of points that walks around, see Fig 3

Additional extensions to the random walk can be easily incorporated, such as bias to a central place:
$$ \mu_t = \mu_{t-1} + \boldsymbol B(a-\mu_{t-1}) + \epsilon_t $$
where $\boldsymbol B$ is a matrix that governs the degree of attraction to point a.

Adding correlation in direction between the steps creates longer, more realistic move similar to a non central place forager searching through a landscape. 
More complex and modern movement models include multiple states, each with their own set of movement parameters, which can be easily done with a piecewise equation for $n$ states.
Movement parameters can be a function of covariates such as age and sex. 

Potential functions allow for the modeling of movement along environmental gradients, the gradient is a vector field of partial derivatives that convey the direction and magnitude of attraction. 

Limiting Distributions<br>
When going from Lagrangian vs Eulerian, the limiting distribution of the time series process is a useful concept, which is the limit of $$ \mu_{t+h} | \mu_t $$ as h -> $\infty$

Conceptually, this is creating greater and greater time gaps between locations, so that they are independent. This limiting distribution of $N$ individuals in $M$ state space is an animal density surface, and is thus conceptually united to SCR model parameters. 

Few movement models have closed form limiting distributions, or limiting distributions at all, but the biased random walk does, see Eqn 11. 

The SCR observation model can be a non-uniform thinning of the limiting distribution of the movement of all individuals in $M$, this is most realistic when an animal can move throughout its entire home range within an occasion. 
The Langevin diffusion RSF model also has a closed form limiting distribution. 

Movement data has GPS error, and is thus often modeled hierarchically with a movement and measurement sub model. 

The scale parameter assumptions have led SCR to be most widely used with species with symmetric and resident home ranges, as opposed to nomadic species. 
Uncertainty can be underestimated, even though abundance is not biased when the modeled species has a non-circular home range. 
Considerable bias can occur with sigma (think dune example from Taal) as observations that are very far away assume a large circle of higher detection probability. 
Resource selection can bias density becuase detection is not just a single function of distance. 
Movement also becomes extremely important in open population model parameter estimation. 
Animal movement models are not typically attached to demographic rates or population estimates. 

Integrating SCR and Movement Data<br>
Basic integrated SCR model can be described as
$$[y|\mu, N, \theta] = [y|\mu, \theta ] [\mu | \theta] [\mu_1, N, |\theta] $$
Term 1 is SCR. 
Term 2 is the movement model.
Term 3 is the point process model.
$\mu$ the position process, is now a latent variable, as we do not know the exact position of the animal in each time step. 
Thus, we estimate a latent movement trajectory as opposed to a static activity center.
Therefore, the distance for the detection kernel becomes the distance between the detector and the latent location at time $t$ as opposed to the activity center $s$. 
The point process now represents the initial positon of N individuals as opposed to N activity centers.
By adding a center of attraction, one can estimate second and third order resource selection (selection of a home range within $M$ and selection of resources within a home range).

One issue with uniting movement data and SCR data is a discrete time versus continuous time mismatch.
One solution is to model the distance between the expected location of the individual within the time step and the observed location, this seems conceptually similar to a dynamic activity center (it is currently unclear to me how it is not the same thing).

Next Steps <br>
One challenge with movement-scr integrated models is optimal sampling is unknown. 
In SCR, the sampling trade of is between the number of individuals captured and the extent of spatial recaptures, which is often dictated by trap spacing and is typically suggested to be $1-2\sigma$. 
Without auxiliary location data, the number of spatial recaptures required is likely to be higher in order to more appropriately model the movement process.

Another challenge is that sampling bias is controlled in SCR through the detection process, but sampling bias is typically not corrected for in movement studies, with telemetered individuals not representing a random sample of the population, and thus potentially biasing the parameters of the movement model from the true population mean.

Models can be broken down into 4 quadrants, space and time bring either discrete or continous. 

Movement near the boundaries of the state space can be a challenge, as simple random walk difffusive movement would lead to a deterministic reduction in density over time that is unrealistic.
A simple solution is expanding the state space far beyond the boundaries of the movement of individuals, which can lead to longer model fitting processes due to the need to estimate more parameters. 
Additionally, now that each movement trajectory is latent as opposed to just the home range center, there are $NT$ latent parameters to estimate as opposed to simply $N$. 
